---
title: "Evaluate reproducibility of Hi-C data with `hicrep`"
author: "Tao Yang  <xadmyangt@gmail.com>"
date: "2/24/2020"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Evaluate reproducibility of Hi-C data with `hicrep`}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

##Introduction

Hi-C data analysis and interpretation are still in their early stages. 
In particular, there has been a lack of sound statistical metric to 
evaluate the quality of Hi-C data. When biological replicates are not 
available, investigators often rely on eithervisual inspection of Hi-C
interaction heatmap or examining the ratio of long-range interaction 
read pairs over the total sequenced reads, neither of which are supported
by robust statistics. When two or more biological replicates are available,
it is a common practice to compute either Pearson or Spearman correlation 
coefficients between the two Hi-C data matrices and use them as a metric 
for quality control. However, these kind of over-simplified approaches are
problematic and may lead to wrong conclusions, because they do not take 
into consideration of the unique characteristics of Hi-C data, such as 
distance-dependence and domain structures. As a result, two un-related 
biological samples can have a strong Pearson correlation coefficient, while
two visually similar replicates can have poor Spearman correlation coefficient.
It is also not uncommon to observe higher Pearson and Spearman correlations 
between unrelated samples than those between real biological replicates. 

we develop a novel framework, `hicrep`, for assessing the reproducibility of 
Hi-C data. It first minimizes the effect of noise and biases by smoothing 
Hi-C matrix, and then addresses the distance-dependence effect by stratifying
Hi-C data according to their genomic distance. We further adopt a 
stratum-adjusted correlation coefficient (SCC) as the measurement of Hi-C data
reproducibility. The value of SCC ranges from -1 to 1, and it can be used to 
compare the degrees of differences in reproducibility. Our framework can also
infer confidence intervals for SCC, and further estimate the statistical 
significance of the difference in reproducibility measurement for different 
data sets. 


In this Vignette, we explain the method rationale, and provide guidance to 
use the functions of `hicrep` to assess the reproducibility for Hi-C 
intrachromosome replicates. 

## Citation

Cite our paper:

HiCRep: assessing the reproducibility of Hi-C data using a 
stratum-adjusted correlation coefficient. Tao Yang, Feipeng Zhang, Galip
Gurkan Yardimci, Fan Song, Ross C Hardison, William Stafford Noble, 
Feng Yue, Qunhua Li. Genome Research 2017. doi: 10.1101/gr.220640.117.



## Installation

Download the source package (hicrep_xxx.tar.gz) from Github.
Or install it from Bioconductor:
```
## try http:// if https:// URLs are not supported
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("hicrep")
```

## Rationale of method

This is a 2-step method (Figure1). In Hi-C data it is often difficult to 
achieve sufficient coverage. When samples are not sufficiently sequenced, 
the local variation introduced by under-sampling can make it difficult to 
capture large domain structures. To reduce local variation, we first smooth 
the contact map before assessing reproducibility. Although a smoothing filter
will reduce the individual spatial resolution, it can improve the contiguity
of the regions with elevated interaction, consequently enhancing the domain 
structures. We use a 2D moving window average filter to smooth the Hi-C 
contact map. This choice is made for the simplicity and fast computation of 
mean filter, and the rectangular shape of Hi-C compartments.

In the second step, we stratify the Hi-C reads by the distance of contacting
loci, calculate the Pearson correlations within each stratum, and then 
summarize the stratum-specific correlation coefficients into an aggregated 
statistic. We name it as Stratum-adjusted Correlation Coefficient (SCC). 
For the methodology details, please refer to our paper on Genome Research.

![Figure1. `hicrep` pipeline schematic representation]
(https://raw.githubusercontent.com/MonkeyLB/hicrep/master/vignettes/hicrep-pipeline.JPG)

```{r load_packages, include=FALSE}
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
library(hicrep)
data("HiCR1")
data("HiCR2")
```


## The format of input and Pre-processing

The input are two Hi-C matrices to be compared.  The Hi-C matrices 
should have the dimension $N\times N$. 

```{r}
dim(HiCR1)
HiCR1[1:10,1:10]
```

The function `get.scc` will first smooth the HiC matrix, with given 
neighborhood size parameter $h$, and filter the bins that have 
zero counts in both replicates. The arguments includes the two matrices, 
the resolution of matrices, smoothing parameter, and the lower bound and 
upper bound of interaction distance considered. The resolution is simply
the bin size. Smoothing parameter decides the neighborhood size of 
smoothing. Below (Figure 2) is a representation of smoothing neighborhood
for a point $C_{ij}$:

![Figure2. A smoothing neighborhood with $h = 2$]
(https://raw.githubusercontent.com/MonkeyLB/hicrep/master/vignettes/Smoothing_neighborhood.JPG)


## Calculate Stratum-adjusted Correlation Coefficient (SCC)

An example to calculate SCC for a matrix of 1Mb resolution. Smoothing 
parameter $h$ is set to 2. The lower bound of distance considered is 0 
(diagnal), and the upper bound is 5Mb.

```{r, eval=TRUE}
scc.out = get.scc(HiCR1, HiCR2, 1000000, 2, 0, 5000000)

#SCC score
scc.out$scc
#Standard deviation of SCC
scc.out$std
```

The output is a list of results including stratum specific Pearson 
correlations, weight coefficient, SCC and the asymptotic standard 
deviation of SCC. The last two numbers are the ones we needed in most
of the situations. 

## Smooth the Hi-C matrix with 2D mean filter

The function `fast.mean.filter()` is a very fast algorithm that applies 
2D mean filter to squred matrices such as Hi-C contact mapes. The output 
is a smoothed matrix that has the same size with the original matrix. 
Here is an example to smooth the matrix with parameter $h = 2$:

```{r, eval = TRUE}

smd_mat = fast.mean.filter(HiCR1, 2)

```

## Select the optimal smoothing parameter

To select $h$ objectively, we develop a heuristic procedure to search 
for the optimal smoothing parameter.  Our procedure is designed based 
on the observation that the correlation between contact maps of replicate 
samples first increases with the level of smoothness and plateaus when 
sufficient smoothness is reached.To proceed, we use a pair of reasonably 
deeply sequenced interaction map as the training data. We randomly 
sampled 10% of the data, then compute SCC for the sampled dataeach 
fraction at a series of smoothing parameters in the ascending order. We
choose the smallest $h$ at which the increment of the average 
reproducibility score is less than 0.01. This procedure is repeated ten
times, and the mode among the ten $h$’s is picked.

```{r, eval=TRUE}
h_hat <- htrain(HiCR1, HiCR2, 1000000, lbr = 0, ubr = 5000000, range = 0:2)

h_hat
```

![Figure3. Select the optimal smoothing parameter $h$]
(https://raw.githubusercontent.com/MonkeyLB/hicrep/master/vignettes/Smoothing_parameter.JPG)

The above graph shows the change of SCC as the $h$ increases from 0 to 10 
for a 40Kb resolution matrix. The parameter $h = 5$ is selected as the 
optimal smoothing neighborhood size.

Important note:
The smoothing parameter selection could be confounded by the sequencing 
depth. Insufficient sequencing depth data might lead to inflated 
smoothing neighborhood size. To compare SCC between pairs of 
replicates that has the same resolution, one shall use the same 
smoothing parameter.

Train the smoothing parameter could be time-consuming. It is not 
suggested to train $h$ every time when calculating SCC. For a giving 
resolution, one could use a deeply sequenced biological replicates 
to train $h$ (i.e., > 300 million total nubmer of reads for whole 
chromosome), and use the trained $h$ for other same resolution data. 
Here we provide some trained $h$ trained based on two replicates 
of hESC cells from Dixon et al 2015 (GEO accession: GSE52457): 

Resolution ($h$): 
10K (20), 25K (10), 40k(5), 100k(3), 500k(1 or 2), 1M(0 or 1). 

## Equalize the total number of reads

In previous section, we mention that sequencing depth could be a confounding
effect. If the total numbers of reads are very different between the two 
replicates, it's suggested that one should down-sample the higher sequencing 
depth to make it equal to the lower one. The best way to do it is to use the 
bam files to do the sub-sampling randomly. In case you only have the matrix 
file available, we made a function `depth.adj()` to do down-sampling from 
matrix files.

```{r}
#check total number of reads before adjustment
sum(HiCR1)

# sub-sample 200000 total reads
DS_HiCR1 <- depth.adj(HiCR1, 200000) 

#check total number of reads after adjustment
sum(DS_HiCR1)
```

```{r}
sessionInfo()
```

```{r}
#check total number of reads before adjustment
sum(HiCR1)

# sub-sample 200000 total reads
DS_HiCR1 <- depth.adj(HiCR1, 200000) 

#check total number of reads after adjustment
sum(DS_HiCR1)
```

## Computation efficiency

Given a pair of contact maps of human chromosome 1 with bin-size equal to 
40kb, it takes 27 seconds on a laptop with 2.6GHz Intel Core i7-6600U and 
16Gb of RAM.
